
```{r setup, include=FALSE, echo=FALSE}
require(knitr)
knitr::opts_chunk$set(eval=TRUE, results=TRUE, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.height=5, fig.width=8, fig.path="figures/")
# knitr::opts_knit$set(root.dir="R/Rinstall")
```

# (PART) Implementation {-}

# R 


## Setup

The main library we will use is `stats` and comes bundled with base R. However, we also need to install a few additional libraries onto our machine and then load them into our search path. 

```{r}
# uncomment the line below to install the {pacman} library on your computer
# install.packages("pacman")
pacman::p_load(
  lme4,         # model specification / estimation
  lmerTest,     # provides p-values in the model output
  future,       # parallelization
  future.apply, # fast automation
  furrr,        # fast functional programming
  faux,         # simulate from multivariate normal distribution
  broom.mixed,  # extracting tidy data from model fits
  tidyverse,    # data wrangling and visualisation
  gt            # nice tables
  )
```

We will also set the pseudo-random number generator seed to 02138 to make the stochastic components of our simulations reproducible. 

```{r}
set.seed(02138)
```

Finally, let's take advantage of background parallelization to speed-up R processes.

```{r}
plan(multisession)
```

## Data simulation step by step

Let's write some code that creates datasets under the alternative hypothesis.

### Establish the data-generating parameters

```{r params-all}
# set all data-generating parameters
beta_0  <-  65   # intercept; i.e., the grand mean
beta_1  <-  15   # slope; i.e, effect of category
tau_0   <-   7   # by-subject random intercept sd
tau_1   <-   4   # by-subject random slope sd
rho     <-   0.2 # correlation between intercept and slope
sigma   <-   8   # residual (error) sd
```

### Simulate the sampling process

```{r sampling}
# set number of subjects and items
n_subj <- 100 # number of subjects
n_pop  <-  25 # number of songs in pop category
n_rock <-  25 # number of songs in rock category
```

### Simulate the sampling of songs

```{r}
# simulate a sample of songs
songs <- tibble(
  song_id = seq_len(n_pop + n_rock),
  category = rep(c("pop", "rock"), c(n_pop, n_rock)),
  genre_i = rep(c(0, 1), c(n_pop, n_rock))
)
```

### Simulate the sampling of subjects

We will use the function `faux::rnorm_multi()`, which generates a table of `n` simulated values from a multivariate normal distribution by specifying the means (`mu`) and standard deviations (`sd`) of each variable, plus the correlations (`r`), which can be either a single value (applied to all pairs), a correlation matrix, or a vector of the values in the upper right triangle of the correlation matrix.

```{r}
# simulate a sample of subjects

# sample from a multivariate normal distribution
subjects <- faux::rnorm_multi(
    n = n_subj,
    mu = 0, # means for random effects are always 0
    sd = c(tau_0, tau_1), # set SDs
    r = rho, # set correlation, see ?rnorm_multi
    varnames = c("b_0j", "b_1j")
  ) |>
  mutate(subj_id = seq_len(n_subj)) # add subject IDs
```

### Check the simulated values

```{r}
tibble(
  parameter = c("tau_0", "tau_1", "rho"),
  value = c(tau_0, tau_1, rho),
  simulated = c(
    sd(subjects$b_0j),
    sd(subjects$b_1j),
    cor(subjects$b_0j, subjects$b_1j)
  )
)
```

### Simulate trials

```{r}
# cross subject and song IDs; add an error term
trials <- crossing(subjects, songs) |>
  mutate(e_ij = rnorm(n(), mean = 0, sd = sigma))
```

### Calculate response values

```{r}
dat_sim <- trials |>
  mutate(liking_ij = beta_0 + b_0j + (beta_1 + b_1j) * genre_i + e_ij) %>%
  select(subj_id, category, genre_i, liking_ij)
```

### Plot the data

```{r}
dat_sim |>
ggplot(aes(category, liking_ij, color = category)) +
  # predicted means
  geom_hline(yintercept = (beta_0 + 0*beta_1), 
             color = "orange", linetype = "dashed", linewidth = 1) +
  geom_hline(yintercept = (beta_0 + 1*beta_1), 
             color = "dodgerblue", linetype = "dashed", linewidth = 1) +
  # actual data
  geom_violin(alpha = 0.5, show.legend = FALSE, fill = "grey65") +
  stat_summary(fun = mean, geom="crossbar", show.legend = FALSE) +
  scale_color_manual(values = c("orange", "dodgerblue")) +
  ggtitle("Predicted versus simulated values") +
  theme_bw()
```

### Analyze the simulated data

```{r}
# fit a linear mixed-effects model to data
mod_sim <- lmer(liking_ij ~ 1 + genre_i + (1 + genre_i | subj_id), data = dat_sim)

summary(mod_sim, corr = FALSE)
```

Use `broom.mixed::tidy(mod_sim)` to get a tidy table of the results. Below, we added column "parameter" and "value", so you can compare the estimate from the model to the parameters you used to simulate the data. 

```{r}
# get a tidy table of results
broom.mixed::tidy(mod_sim) |>
  mutate_if(is.numeric, round, 3) |>
  mutate(
    parameter = c("beta_0", "beta_1", "tau_0", "rho", "tau_1", "sigma"),
    value = c(beta_0, beta_1, tau_0, rho, tau_1, sigma),
  ) |>
  select(term, parameter, value, estimate) |>
  knitr::kable()
```

## Data simulation automated

Now that we've tested the data generating code, we can put it into a function so that it's easy to run it repeatedly.

```{r}
# set up the custom data simulation function
sim_data <- function(
  n_subj     = 100,   # number of subjects
  n_pop      =  25,   # number of pop songs
  n_rock     =  25,   # number of rock songs
  beta_0     =  65,   # mean for pop genre
  beta_1     =  15,   # effect of genre
  tau_0      =   7,   # by-subject random intercept sd
  tau_1      =   4,   # by-subject random slope sd
  rho        =   0.2, # correlation between intercept and slope
  sigma      =   8    # residual (standard deviation)
  )
{
  # simulate a sample of songs
  songs <- tibble(
    song_id = seq_len(n_pop + n_rock),
    category = rep(c("pop", "rock"), c(n_pop, n_rock)),
    genre_i = rep(c(0, 1), c(n_pop, n_rock))
  )

  # simulate a sample of subjects
  subjects <- faux::rnorm_multi(
    n = n_subj,
    mu = 0,
    sd = c(tau_0, tau_1),
    r = rho,
    varnames = c("b_0j", "b_1j")
  ) |>
  mutate(subj_id = seq_len(n_subj))

# cross subject and song IDs
crossing(subjects, songs) |>
  mutate(e_ij = rnorm(n(), mean = 0, sd = sigma),
         liking_ij = beta_0 + b_0j + (beta_1 + b_1j) * genre_i + e_ij) |>
  select(subj_id, category, genre_i, liking_ij)
}
```

## Power calculation (single run)

We can wrap the data generating function and analysis code in a new function (`power_run()`) that returns a tidy table of the analysis results.

```{r}
# set up the power function
power_run <- function(...) {
  # ... is a shortcut that forwards any additional arguments to sim_data()
  dat_sim <- sim_data(...)
  mod_sim <- suppressWarnings({ suppressMessages({
    lmerTest::lmer(liking_ij ~ 1 + genre_i + (1 + genre_i | subj_id), data = dat_sim)
  })})
  broom.mixed::tidy(mod_sim)
}
```

```{r}
# run one model with default parameters
power_run()
```

```{r}
# run one model with new parameters
power_run(n_pop = 50, n_rock = 25, beta_1 = 5)
```

## Power calculation automated

To get an accurate estimation of power, we need to run the simulation many times. We use 100 here as an example, but the results will be more accurate the more replications you run. This will depend on the specifics of your analysis, but we recommend at least 1000 replications.

```{r}
reps <- 100
sims <- map_df(1:reps, ~ power_run())
```

```{r}
# calculate mean estimates and power for specified alpha
alpha <- 0.05

sims |>
  filter(term == "genre_i") |>
  group_by(term) |>
  summarise(
    mean_estimate = mean(estimate),
    mean_se = mean(std.error),
    power = mean(p.value < alpha),
    .groups = "drop"
  )
```

### Check false positive rate

Set the effect of `genre_ij` to 0 to calculate the false positive rate. This is the probability of concluding there is an effect when there is no actual effect in your population.

```{r}
# run simulations and calculate the false positive rate
reps <- 100
sims_fp <- map_df(1:reps, ~ power_run(beta_1 = 0))

# calculate mean estimates and power for specified alpha
alpha <- 0.05

sims_fp |>
  filter(term == "genre_i") |>
  summarise(power = mean(p.value < alpha))
```

Ideally, the false positive rate will be equal to alpha, which we set here at `r alpha`. 

## Power for different effect sizes

In real life, we will not know the effect size of our quantity of interest and so we will need to repeatedly perform the power analysis over a range of different plausible effect sizes.

```{r}
# grid of paramater values
pgrid <- crossing(
  n_subj = c(10, 25, 50, 100),
  n_pop = c(10, 50),
  n_rock = c(10, 50),
  beta_1 = 1:5 
)

# fit the models over the parameters
parameter_search <- function(params = pgrid){
  future_pmap_dfr(
    .l = params,
    .f = ~ power_run(n_subj = ..1, 
                     n_pop  = ..2, 
                     n_rock = ..3,
                     beta_1 = ..4),
    .options = furrr_options(seed = TRUE),
    .progress = TRUE
  )
}
```

Fair warning, this will take some time!

```{r}
# replicate parameter search 100 times
reps <- 100

# replicate each row of the parameter grid to match the dimensions of the model outputs
pgrid_expand <- pgrid |> 
  slice(rep(1:n(), each = 6)) |> # expand by 6 parameters
  map_df(rep.int, times = reps) # expand by number of reps

sims_params <- future_replicate(
    n = reps, 
    expr = parameter_search(), 
    simplify = FALSE
    ) |>
  imap( ~ mutate(.x, rep = .y, .before = "effect")) |>
  bind_rows() |>
  mutate(pgrid_expand, .before = "effect") # add in the parameter grid
```

```{r}
# calculate mean estimates and power for specified alpha
alpha <- 0.05

sims_table <- sims_params |>
  filter(term == "genre_i") |>
  group_by(term, n_subj, n_pop, n_rock, beta_1) |>
  summarise(
    mean_estimate = mean(estimate),
    mean_se = mean(std.error),
    power = mean(p.value < alpha),
    .groups = "drop"
  )
```

Here's the table from the simulation.

```{r}
sims_table |>
  gt() |>
  tab_header(title = "Power values for GLM") |>
  data_color(
    columns = power,
    colors = scales::col_numeric(
      palette = c("red", "green"),
      domain = c(0, 1)
      )
  )
```

Here's the graph from the simulation. 

```{r}
sims_table |>
  mutate(across(n_subj:beta_1, as.factor)) |>
  ggplot(aes(x = mean_estimate, y = power,
             group = n_subj, color = n_subj)) +
  geom_line() +
  geom_point(size = 2) +
  facet_grid(n_pop ~ n_rock) +
  ylim(0, 1) +
  theme_bw()
```
