
```{r, include=FALSE, echo=FALSE}
require(knitr)
knitr::opts_chunk$set(eval=TRUE, results=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.height=5, fig.width=8, fig.path="figures/")
# knitr::opts_knit$set(root.dir="R/Rinstall")
```

# (PART) Implementation {-}

# R 

## Simple linear regression

### Setup

The main library we will use is `stats` and comes bundled with base R. However, we also need to install a few additional libraries onto our machine and then load them into our search path. 

```{r}
# install.packages("pacman")
pacman::p_load(lmtest, lme4, dplyr, ggplot2, gt)
```

We will also set the pseudo-random number generator seed to 02138 to make the stochastic components of our simulations reproducible. 

```{r}
set.seed(02138)
```

### Steps 4-5: Simulate and automate

Next, let's write a function that creates datasets under the alternative hypothesis, fits the models, and uses a likelihood-ratio test to calculate power.

```{r}
sample_n <- c(400, 500, 600, 700)
interact_coef <- c(0.2, 0.25, 0.3, 0.35, 0.4)
repeats <- 1:1000

power_list <- data.frame(sample_n=double(), 
                         interact_coef=double(), 
                         power=double()
                         )[-1, ]

for (s in sample_n) {
  
  for (i in interact_coef){
    
    results <- c()
    
    for (r in repeats){
      
      age <- ceiling(runif(s, 18, 65))
      female <- rbinom(s, 1, 0.5)
      interact <- age * female
      e <- rnorm(s, 0, 20)
      sbp <- 110 + 0.5*age + (-20)*female + i*interact + e
      dataset <- data.frame(sbp, age, female, interact)
      
      full_model <- lm(sbp ~ age + female + interact, data=dataset)
      reduced_model <- lm(sbp ~ age + female, data=dataset)
      prob <- lrtest(full_model, reduced_model)$Pr[2]
      reject <- ifelse((prob<=0.05), 1, 0)
      
      results <- rbind(results, reject)
      
    }
    
    power_list <- rbind(power_list, 
                        data.frame(sample_n=s, 
                                   interact_coef=i, 
                                   power=mean(results)
                                   )
                        )
    
  }  
}
```

### Step 6: Summarize & visualize

Here's the table from the simulation.

```{r}
power_list |>
  gt() |>
  tab_header(title="Power values for GLM") |>
  data_color(
    columns=power,
    colors=scales::col_numeric(
      palette=c("red", "green"),
      domain=c(0, 1)
      )
  )
```

Here's the graph from the simulation. 

```{r}
power_list |>
  mutate(sample_n = factor(sample_n)) |>
  ggplot(aes(x=interact_coef, y=power, 
             group=sample_n, color=sample_n)) + 
  geom_line() + 
  geom_point(size=4) + 
  ylim(0, 1) +
  theme_bw()
```

## Mixed effects model

For the mixed effects model example, we will continue to use the same R libraries and pseudo-random number generator seed as previously.

### Step 4-5: Simulate & automate

Next, let's write a function that creates datasets under the alternative hypothesis, fits the mixed effects models, tests the null hypothesis of interest, and uses a for loop to run many iterations of the function. 

```{r}
sample_n <- c(100, 200, 300, 400, 500)
obs_n <- c(3, 5)
reps_n <- 1:100

power_list_mixed <- data.frame(obs=integer(), 
                               sample=integer(), 
                               power=double()
                               )

for (s in sample_n){
  
  for (o in obs_n){
    
    results <- c()
    
    for (r in reps_n){
      
      data.frame(
        child = factor(1:s), 
        female = rbinom(s, 1, 0.5),
        u_0i = rnorm(s, 0, 0.8),
        u_1i = rnorm(s, 0, 1.7)
        ) ->
      data_set
      
      data_set_expand <- data_set[rep(seq(nrow(data_set)), o), 1:4]

      age <- c()
      for (obser in seq(0, (o-1)*0.5, 0.5)){
        age <- c(age, rep(obser, s))
      }

      data_set_expand |>
        mutate(age = age,
               e_ij = rnorm(s*o, 0, 1.35),
               interact = age * female,
               weight = 5.35 + 2.1*age + (-0.35)*female + (-0.55)*interact + u_0i + age*u_1i + e_ij
               ) ->
      data_set_expand

      full_model <- lmer(weight ~ age + female + interact + (age || child), 
                         data=data_set_expand)
      reduced_model <- lmer(weight ~ age + female + (age || child), 
                            data=data_set_expand)
      prob <- lrtest(full_model, reduced_model)$Pr[2]
      reject <- ifelse((prob<=0.05), 1, 0)
      results <- rbind(results, reject)     
    }
    
    power_list_mixed <- rbind(power_list_mixed, 
                              data.frame(obs=o, 
                                         sample=s, 
                                         power=mean(results)
                                         )
                              )
    
  }
}
```

### Step 6: Summarize & visualize

The table for the mixed effects model.

```{r}
power_list_mixed |>
  gt() |>
  tab_header(title="Power values for mixed effects model") |>
  data_color(
    columns=power,
    colors=scales::col_numeric(
      palette=c("red", "green"),
      domain=c(0, 1)
      )
  )
```

The graph of the simulation result for mixed effects model.

```{r}
power_list_mixed |>
  mutate(obs = factor(obs)) |>
  ggplot(aes(x=sample, y=power, 
             group=obs, color=obs)) + 
  geom_line() + 
  geom_point(size=4) + 
  ylim(0, 1) +
  theme_bw()
```
