# Steps 4-5: Simulate and automate

Next, let's write a function that creates datasets under the alternative hypothesis.

```{r, eval=FALSE}
generate_design_matrix <- function(
  n_participants,
  n_songs,
  genres = c("rock", "pop")
  )
{
  crossing(
    participant = 1:n_participants,
    genre = genres,
    song = 1:n_songs
    ) |>
  mutate(song = paste(genre, song, sep = "_"))
}
```

```{r, eval=FALSE}
data1 <- generate_design_matrix(1000, 2, 1)

group_means <- c(rock = 75, pop = 75) # define means of both genres in a vector
rock_sd <- 7.5 # define sd of rock music
pop_sd <- 7.5 # define sd of pop music
correlation <- 0.2 # define their correlation

sigma <- matrix(c(
  rock_sd^2, rock_sd*pop_sd*correlation,
  rock_sd*pop_sd*correlation, pop_sd^2),
  ncol = 2) # define variance-covariance matrix

set.seed(123)
bivnorm <- data.frame(mvrnorm(nrow(data1)/2, group_means, sigma)) # simulate bivariate normal (we use nrow(data1)/2, the number of rows from the data-set above to simulate 10 observations per group)
par(mfrow=c(1,2))
hist(bivnorm$rock, main = "liking of rock music", xlab = "")
hist(bivnorm$pop, main = "liking of pop music", xlab = "")
```

A matrix containing the variances of both groups (rock and pop) and their covariance (i.e. $\textrm{rock_sd}^2 = 56.25$ and $\textrm{pop_sd}^2 = 56.25$ and the covariances between the groups $\textrm{rock_sd} \times \textrm{pop_sd} \times \textrm{correlation} = 11.25$) that looks like this:

$$
\begin{bmatrix}
56.25 & 11.25 \\
11.25 & 56.25
\end{bmatrix}
$$



```{r, eval=FALSE}
bivnorm_dat <- data.frame(cbind(liking = c(bivnorm$rock, bivnorm$pop), genre = c(rep("rock", (nrow(data1)/2)), rep("pop", (nrow(data1)/2))), participant = rep(1:(nrow(data1)/2), 2))) # this just converts bivnorm to long format and adds the genre variable

bivnorm_dat$liking <- as.numeric(bivnorm_dat$liking) # variable was converted to character for some reason in cbind so lets make it numerical again

bivnorm_dat$genre_f <- factor(bivnorm_dat$genre) # make genre a factor

lmer_bivnorm <- lmer(liking ~ genre_f + (1 | participant), bivnorm_dat)
summary(lmer_bivnorm)
```


Next, let's write a function that creates datasets under the alternative hypothesis, fits the models, and uses a likelihood-ratio test to calculate power.